A game-theory modeling approach to utility and strength of interactions dynamics in biomedical research social networks

### 2017-03-30 casm.tex
acknowledge copy editing
Added new references as requested by reviewer 1. Also, marked in red additions and modifications to the text. New references in the .bib file.

Nowadays it is widely acknowledged that collaboration takes many
forms, from sharing of biological samples and biobanking to
international groups in charge of helping research communities to
harmonize and share their data. Sharing resources such as equipment,
funds, and time is critical; building trust among scientists is
fundamental. Also, resources are mobilized in order to create
strategic alliances.\\

The analysis of cooperation in scientific research has been the
subject of a number of studies
\cite{VermeulenPenders:2013,Newman:2001,Newman:2004,Elango-etal:2012,HernandezLemus:2013,Strasser:2006,Strasser:2012}. This
is not surprising since cooperation and competition are quite
important in today's academic success. Moreover, the quantitative
analysis of cooperation in science also has a long history of more
than 50 years
\cite{Mali-etal:2012,Scharnhorst-etal:2012}. Particularly appealing to
researchers studying this topic is the use of network analysis,
especially in the area of bibliometrics and scientometrics such as in
citation networks that date as back as 1972 \cite{Garfield:1972},
co-authorship and collaboration networks also have a long tradition
\cite{Todorov-etal:1991,Wagner-etal:2017}.\\

The approach we follow here articulates game theory together with
complex network analysis. Putting together game-theoretic modeling on
complex networks is common in studies about the emergence of
cooperation
\cite{SzaboFath:2007,Nowak-etal:92,OshtukiNowak:2006,Santos-etal:2005,Santos-etal:2006}.
However, in the context of science studies, it is far from being a
unified, active field of research. There is no common set of questions
and problems or a coherent body of models and theory aimed to
understand the quirks and twists of cooperation in scientific
communities. It is more likely to say that work of this kind is
scarce and their implementatoin is on different situations related to
the social life of science, where sometimes models are run on science
collaboration networks because access to these kind of data is
relatively easy to access.\\

Nevertheless, there is work done. For instance Wardil and Hauert have
analyzed how does collaboration happen within co-authorship networks
\cite{Wardil-etal:2015}. Using the \emph{Physical Review} journal
historic database, these authors show the payoff differences between
being the first author and the last author, where the first author is
a cooperator that takes the highest load of work, while the last
author is the \emph{laird} that can benefit from the work of the
former and get engaged in many other projects. They argue that
cooperation in a multi-authored paper resambles more to a snow-drift
game than a prisoner's dilemma situation, since when co-authoring a
paper, every body wins but each author pays a different price. Though
this work takes into account the topology of the network as
informative regarding cooperation patterns, the role of topology
framing or fostering cooperation is not part of the study. \\

Game theory over complex scientific information and collaboration
networks has also attracted attention. Hanauske has applied
\emph{Evolutionary Game Theory} to have a better understanding of the
market of scientific information, that involves scientists as
producers but also journal, that are in charge of its
dissemination. One of his models connects scientist interest in
maximazing their reputation and their decissions to publish in an Open
Access Journal or in a journal that works under the traditional
publication framework. He has also modeled a game-theoretic
interactions between scientists and journals for both players to go
open access or not. Only this second model incorporates networks
analysis \cite{Hanauske-etal:2012}.\\

Another interesting behavior explored by game-theoretical modeling and
complex networks has been done by Li and Cao \cite{Li-etal:2009}. This
authors analyze the role of punishment in a science collaboration
network. More precisely, they implemented an ultimatum bargaining
model to understand the impact of the topology of a social network and
the emergence of a population of punishers. In the ultimatum game,
there are two players, one of them (the proponent) suggest to split a
sum in two and its connected neighbor (the respondent), can decide to
accept or reject the offer as it may be perceived more or less
fair. If the respondent accepts, the sum is divided according to the
offer. If the offer is rejected both players do not get
anything. According to the authors, the act of rejecting a selfish
offer can be interpreted as a punishment since the player that rejects
it is paying a cost of not receiving anything but the proponent looses
even more, since his offer was selfish. Their results show that the
population of punishers will emerge as a phase transition as the
degree of selfishness increses among players. Moreover, they found a
dependence relation between the tolerance to selfish behaviors and the
topologies of networks.\\

@@

In a similar vein to the work of Li and Cao, we explore the network
effect on the distribution of players having access to certain amount
of resources shared by other players in the network and the
distribution of the strength of interactions among them. Particularly,
we implemented two games played simultaneously: one for maximizing
individual utility based on the iterated Prisoner's Dilemma; the
other, a coordination game for maximizing the connection strength
between players. We are interested in how they affect each other in
the context of a network of scientific collaboration under the idea
that while researchers are interested in maximizing their utilities,
they also know that it is important to invest in building
collaborative relationships. These two behaviors are explored in a
biomedical research community in M\'exico.\\

@@

In the year 2002 CONACyT, along with other government agencies and
entities, created sectoral funds
to cover and equally
promote research capacities of different areas such as energy, agriculture
and health. Technological innovation is fostered by the generation of human resources
and by helping research groups to consolidate. It is expected that the
knowledge generated under the sponsorship of these funds
will be the product of applied research that attends national public
needs, and promotes economic growth.\\

\emph{FOSISS} or Sectoral Fund for Health and Social Security Research
(\emph{Fondo Sectorial en Investigaci\'on en Salud y Seguridad Social}) is one of
such funds. FOSISS is constituted by CONACyT, SSA, IMSS and ISSSTE,\footnote{SSA
  is the acronym for Department of Health (\emph{Secretar\'ia de Salud}); IMSS is
  the acronym for Social Security Mexican Institute (\emph{Instituto Mexicano
    del Seguro Social}); ISSSTE stands for Institute for Social Security and
  Services for State Workers (\emph{Instituto de Seguridad y Servicios Sociales
    de los Trabajadores del Estado})} all of them being the major public health
providers and research institutions in the country. Every year CONACyT opens a
call for funds limited to a set of health research areas previously defined by a
group of experts. Most applicants are public universities and research
institutions, but eligibility is open to public and private health research sectors.
From 2002 through 2013, there were 91 institutions funded that
comprised 4,988 researchers.\\

@@

Besides the characteristics of the population there are some other boundary
conditions that play an important role on the network topology and dynamics
that motivated the development of our model. Biomedical research in M\'exico
constitutes a vibrant community and collaboration is part of everyday
work. However, M\'exico does not have public biobanks for research purposes
(which are especially relevant for research in genomics, for example), there is
no regulation on the access to biological samples such as tissue, cells, DNA,
RNA, etc.\footnote{Regulation exists regarding researcher-subject relations
  based on legal and ethical grounds. Also, all projects need to be approved by
  the Ethics Committee and IRB.} Something similar happens with data. There have
been some attempts to create open data repositories for biomedical research, but
they have not been established yet. Regulation on these subjects is still
missing. Finally, technologically advanced equipment such as high throughput sequencers
are kept by institutions with the highest research
profiles and sometimes PIs manage them in a self serving way.\\

@@

\section*{Acknowledgments}
    We acknowledge the help of Ma. del Pilar Galarza and Francisco Allende for their work curating 
databases of researchers. We acknowledge Be Luecke for her critical review.

>>>>

Nowadays it is widely acknowledged that collaboration takes many forms, from
sharing of biological samples and biobanking to international groups in charge
of helping research communities to harmonize and share their data {\color{red}
\cite{VermeulenPenders:2013,Barrett-etal:2013,Federer-etal:2015,Karimi-Busheri:2015,Vallance-etal:2016,Warner-etal:2017}. Sharing
resources such as equipment, funds, and time is critical; building trust among
scientists is fundamental \cite{Bennett-etal:2010,Leite:2017}. Also, resources are mobilized
in order to create strategic alliances.}\\

{\color{red}The analysis of cooperation in scientific research has been the
subject of a number of studies
\cite{VermeulenPenders:2013,Newman:2001,Newman:2004,Elango-etal:2012,HernandezLemus:2013,Strasser:2006,Strasser:2012}. This
is not surprising since cooperation and competition are quite
important in today's academic success. Moreover, the quantitative
analysis of cooperation in science also has a long history of more
than 50 years
\cite{Mali-etal:2012,Scharnhorst-etal:2012}. Particularly appealing to
researchers studying this topic is the use of network analysis,
especially in the area of bibliometrics and scientometrics such as in
citation networks that date as back as 1972 \cite{Garfield:1972},
co-authorship and collaboration networks also have a long tradition
\cite{Todorov-etal:1991,Wagner-etal:2017}.}\\

{\color{red}The approach we follow here articulates game theory together with
complex network analysis. Putting together game-theoretic modeling on
complex networks is common in studies about the emergence of
cooperation
\cite{SzaboFath:2007,Nowak-etal:92,OshtukiNowak:2006,Santos-etal:2005,Santos-etal:2006}.
However, in the context of science studies, it is far from being a
unified, active field of research. There is no common set of questions
and problems or a coherent body of models and theory aimed to
understand the quirks and twists of cooperation in scientific
communities. It is more likely to say that work of this kind is
scarce and their implementatoin is on different situations related to
the social life of science, where sometimes models are run on science
collaboration networks because access to these kind of data is
relatively easy to access.}\\

{\color{red}Nevertheless, there is work done. For instance Wardil and Hauert have
analyzed how does collaboration happen within co-authorship networks
\cite{Wardil-etal:2015}. Using the \emph{Physical Review} journal
historic database, these authors show the payoff differences between
being the first author and the last author, where the first author is
a cooperator that takes the highest load of work, while the last
author is the \emph{laird} that can benefit from the work of the
former and get engaged in many other projects. They argue that
cooperation in a multi-authored paper resambles more to a snow-drift
game than a prisoner's dilemma situation, since when co-authoring a
paper, every body wins but each author pays a different price. Though
this work takes into account the topology of the network as
informative regarding cooperation patterns, the role of topology
framing or fostering cooperation is not part of the study.} \\

{\color{red}Game theory over complex scientific information and collaboration
networks has also attracted attention. Hanauske has applied
\emph{Evolutionary Game Theory} to have a better understanding of the
market of scientific information, that involves scientists as
producers but also journal, that are in charge of its
dissemination. One of his models connects scientist interest in
maximazing their reputation and their decissions to publish in an Open
Access Journal or in a journal that works under the traditional
publication framework. He has also modeled a game-theoretic
interactions between scientists and journals for both players to go
open access or not. Only this second model incorporates networks
analysis \cite{Hanauske-etal:2012}.}\\

{\color{red}Another interesting behavior explored by game-theoretical modeling and
complex networks has been done by Li and Cao \cite{Li-etal:2009}. This
authors analyze the role of punishment in a science collaboration
network. More precisely, they implemented an ultimatum bargaining
model to understand the impact of the topology of a social network and
the emergence of a population of punishers. In the ultimatum game,
there are two players, one of them (the proponent) suggest to split a
sum in two and its connected neighbor (the respondent), can decide to
accept or reject the offer as it may be perceived more or less
fair. If the respondent accepts, the sum is divided according to the
offer. If the offer is rejected both players do not get
anything. According to the authors, the act of rejecting a selfish
offer can be interpreted as a punishment since the player that rejects
it is paying a cost of not receiving anything but the proponent looses
even more, since his offer was selfish. Their results show that the
population of punishers will emerge as a phase transition as the
degree of selfishness increses among players. Moreover, they found a
dependence relation between the tolerance to selfish behaviors and the
topologies of networks.}\\

@@

{\color{red}In a similar vein to the work of Li and Cao,} we explore the network
effect on the distribution of players having access to certain amount
of resources shared by other players in the network and the
distribution of the strength of interactions among them. Particularly,
we implemented two games played simultaneously: one for maximizing
individual utility based on the iterated Prisoner's Dilemma; the
other, a coordination game for maximizing the connection strength
between players. We are interested in how they affect each other in
the context of a network of scientific collaboration under the idea
that while researchers are interested in maximizing their utilities,
they also know that it is important to invest in building
collaborative relationships. These two behaviors are explored in a
biomedical research community in M\'exico.\\

@@

In the year 2002 CONACyT, along with other government agencies and
entities, created sectoral funds
to cover and equally
promote research capacities of different areas such as energy, agriculture
and health. Technological innovation is fostered by the generation of human resources
and by helping research groups to consolidate. It is expected that the
knowledge generated under the sponsorship of these funds
will be the product of applied research that attends national public
needs, and promotes economic growth.  {\color{red} \emph{FOSISS} or Sectoral
Fund for Health and Social Security Research (\emph{Fondo Sectorial en
  Investigaci\'on en Salud y Seguridad Social}) is one of
such funds \cite{Gonzalez-etal:2008}}.\\

FOSISS is constituted by CONACyT, SSA, IMSS and ISSSTE,\footnote{SSA
  is the acronym for Department of Health (\emph{Secretar\'ia de Salud}); IMSS is
  the acronym for Social Security Mexican Institute (\emph{Instituto Mexicano
    del Seguro Social}); ISSSTE stands for Institute for Social Security and
  Services for State Workers (\emph{Instituto de Seguridad y Servicios Sociales
    de los Trabajadores del Estado})} all of them being the major public health
providers and research institutions in the country. Every year CONACyT opens a
call for funds limited to a set of health research areas previously defined by a
group of experts. Most applicants are public universities and research
institutions, but eligibility is open to public and private health research sectors.
From 2002 through 2013, there were 91 institutions funded that
comprised 4,988 researchers.\\

@@

Besides the characteristics of the population there are some other boundary
conditions that play an important role on the network topology and dynamics that
motivated the development of our model. Biomedical research in M\'exico
constitutes a vibrant community and collaboration is part of everyday
work. However, M\'exico does not have public biobanks for research purposes
(which are especially relevant for research in genomics, for example), and the
current regulation on the access to biological samples such as tissue,
cells, DNA, RNA, etc., is very general in scope 
{\color{red}\cite{Motta-Saruwatari:2016}}.\footnote{Regulation exists regarding
  researcher-subject relations based on legal and ethical grounds. Also, all
  projects need to be approved by the Ethics Committee and IRB.} Something
similar happens with data. There have been some attempts to create open data
repositories for biomedical research, but they have not been established
yet. Regulation on these subjects is still missing. Finally, technologically
advanced equipment such as high throughput sequencers are kept by institutions
with the highest research profiles and sometimes PIs manage them in a self serving way.\\

@@

\section*{Acknowledgments}

We acknowledge the help of Ma. del Pilar Galarza and Francisco Allende
for their work curating databases of researchers. We are grateful to
Be Luecke for copy editing the manuscript.
